---
title: "Class 8: Breast cancer mini project"
author: "Shuk Yee Leung (PID: A18050465)"
format: pdf
toc: TRUE
---

## Background

This mini-project explores unsupervised learning techniques applied to the Winsconsin Breast Cancer Diagnostic Data Set, which contains measurements of human breast mass cell nuclei. The project guides the user through exploratory data analysis, performing and interpreting Principal Component Analysis (PCA) to reduce the dimensionality of the data while retaining variance, and applying hierarchical clustering with different linkage methods. The ultimate goal is to combine PCA and clustering to better separate benign and maligment cell samples, evaluating the result using matrics like sensitivity and specificity, and finally demostrating how to predict the classification of new samples using the developed PCA model.

## Data import

Our data come from the U. of Winsconsin Medical Center

```{r}
wisc.df <- read.csv("WisconsinCancer (3).csv", row.names = 1)
```

> Q1. How many patients/samples are in this dataset?

```{r}
nrow(wisc.df)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```
```{r}
sum(wisc.df$diagnosis == "M")
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
colnames(wisc.df)
```
```{r}
length( grep("mean", colnames(wisc.df), value= T) )
```

There is a diagnosis column that is the clinical consenus that I want to exclude from any further analysis. We will come back later and compare our results to this diagnosis.

```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
head(diagnosis)
```

Now we can remove it from the `wisc.df`

```{r}
wisc.data <- wisc.df[,-1]
```

## Clustering

Let's try a `hclus()`
```{r}
hc <- hclust( dist(wisc.data))
plot(hc)
```

We can extract clusters from this rather poor dendrogram/tree with the `cutree()`

```{r}
grps <- cutree(hc, k=2)
```

How many individuals in each clusters?

```{r}
table(grps)
```
```{r}
table(diagnosis)
```

We can generate a cross-table that compares our clusters `grsp` vector with our `diagnosis` venctor values

```{r}
table(diagnosis, grps)
```

## Principal Component Analysis

### The importance of data scaling

The main function for PCA in base R is is `prcomp()`. It has a default input paramater of `scale=FALSE`.

```{r}
#prcomp()
head(mtcars)
```

We could do a PCA of this data as is, and it could be mis-leading...

```{r}
pc <- prcomp(mtcars)
biplot(pc)
```

Let's look at the mean values of each column and their standard deviation. 

```{r}
colMeans(mtcars)
```
```{r}
apply(mtcars, 2, sd)
```

We can "scale" this data before PCA to get a much better representation and analysis of all the columns

```{r}
mtscale <- scale(mtcars)
```

```{r}
round(colMeans(mtscale))
```

```{r}
apply(mtscale, 2, sd)
```

```{r}
pc.scale <- prcomp(mtscale)
```

we can look at the two main results figures from PCA - the "PC plot" (a.k.a score plot, ordienation plot, or PC1 vs PC2 plot). The "loadings plot" how the original variables contribute to the new PCs. 

A loading plot of the unscalled PCA results
```{r}
library(ggplot2)

ggplot(pc$rotation) + 
  aes(PC1, rownames(pc$rotation)) +
  geom_col()

```

Loadings plot of the scalled data
```{r}
ggplot(pc.scale$rotation) +
  aes(PC1, rownames(pc$rotation)) +
  geom_col()
```

PC plot of scaled PCA results
```{r}
library(ggrepel)

ggplot(pc.scale$x) +
  aes(PC1, PC2, label=rownames(pc.scale$x)) +
  geom_point() +
  geom_text_repel()
```

> **Key point**: In general we will set `scale=TRUE` when we do PCA. This is not the default but probably should be...

We can check the SD and mean of the different columns in`wisc.data` to see if we need to scale - hiny: we do!

### PCA of wisc.data

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
```

To see how well PCA is doing here in terms capturing the variance (or speed) in the data we can use the `summary()` function.

```{r}
summary(wisc.pr)
```

Let's make the main PC1 vs PC2
```{r}
ggplot(wisc.pr$x) + 
  aes(PC1, PC2, col=diagnosis) +
  geom_point() +
  xlab("PC1 (44.3%)") +
  ylab("PC2 (19%)")
```
> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.3% (O.4427) of the original variance is captured by PC1.

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs are required to describe at least 70% of the original variance.

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs are required to describe at least 90% of the original variance

>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

The points on the biplot were clustered together. There is 7 unlabeled data points  because there were too many overlaps. It is difficult to understand because of too many data points clustered together ans there is no clear separation of benign and malignant diagnoses groups. 

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(wisc.pr$x[, c(1, 3)], col = diagnosis, 
     xlab = "PC1", ylab = "PC3")
```

>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation [,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
summary(wisc.pr)
```
5 PCs required to explain 80% of the variance of the data

## 5. Combinding methods 

We can take our PCA results and use them as a basis set for other analysis such as clustering.

### Clustering on PCA results

```{r}
wisc.pr.hclust <- hclust( dist( wisc.pr$x[ ,1:2]), method="ward.D2" )
plot(wisc.pr.hclust)
```

We can "cut" this tree to yeild our clusters (groups):
```{r}
pc.grps <- cutree(wisc.pr.hclust, k=2)
table(pc.grps)
```

How do my cluster grps compare to the expert diagnosis

```{r}
table(diagnosis, pc.grps)
```

```{r}
table(diagnosis)
```

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

Ward.D2 because it minimizes variance within clusters to make sure that the cluster are well-separated.

> Q13. How well does the newly created model with four clusters separate out the two diagnoses?

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
# Compare to actual diagnoses
table(wisc.pr.hclust.clusters, diagnosis)
```

> Q14. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
wisc.km <- kmeans(wisc.data, centers = 2)
table(wisc.km$cluster, diagnosis)
```
```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=4)
table(wisc.pr.hclust.clusters, diagnosis)
```

They did really badly. we do much matter after PCA - the new PCA variables (what we call a basis set) give as much better seperation of M and B

## 7. Prediction

We can use our PCA model for the analysis of new "unseen" data. In this case from U. Mich. 

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[, c(1, 2)], col= diagnosis)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

People with malignant cancer (2). 
