---
title: "Class 7: Machine Learning 1"
author: "Shuk Yee Leung (PID: A18050465)"
format: pdf
toc: TRUE
---

Today we will explore unsupervised machine learning methods starting with clustering and dimensionality reduction.

## Clustering

To start, let's make up some data to cluster where we know what the answer should be. The `rnorm()` function will help us here. 

```{r}
hist( rnorm(10000, mean=3) )
```

Return 30 numbers centered on -3

```{r}
tmp <- c( rnorm(30, mean=-3), 
   rnorm(30, mean=+3) )

x <- cbind(x=tmp, y=rev(tmp))

x
```

Make a plot of `x`

```{r}
plot(x)
```

### K-means

The main function in "base" R for K-means clustering is called `kmeans()`:

```{r}
km <- kmeans(x, centers = 2)
km
```

The `kmeans()` function return a "list" with 9 components. You can see the named components of any list with the `attributes()` function.

```{r}
attributes(km)
```

> Q. How many points are in each cluster?

```{r}
km$size
```

> Q. Cluster assignment/membership vector?

```{r}
km$cluster
```

> Q. Cluster centers?

```{r}
km$centers
```

> Q. Make a plot of our `kmeans()` results showing cluster assignment using different colors for each cluster/group of points and cluster centers in blue.

```{r}
plot(x, col=km$cluster )
points(km$centers, col="blue", pch=15, cex=2)
```

> Q. Run `kmeans()` again on `x` and this cluster into 4 groups/clusters and plot the same result figure as above.

```{r}
km4 <- kmeans(x, centers = 4)
km4
```

```{r}
plot(x, col=km4$cluster)
points(km4$centers, col="blue", pch=15, cex=2)
```

> **key-point**: K-means clustering is supper popular but can be miss-used. One big limination is that it can impose a clustering pattern on your data even if clear natural grouping don't exist - i.e. it does what you tell it to so un therms of `centers`. 


### Hierarchical Clustering

The main function in "base" R for Hierarchical clustering is called `hclust()`. 

You can't just pass out dataset as is into `hclust()`, you must give "distance matrix" as input. We can get this from the `dist()` function in R.

```{r}
d <- dist(x)
hc <- hclust(d)
hc
```

The results of `hclust()` don't have a useful `print()` method but do have a special `plot()` method. 

```{r}
plot(hc)
abline(h=8, col="red")
```

To get our main cluster assignment (membershop vector), we need to "cut" the tree at the big goal posts...

```{r}
grps <- cutree(hc, h=8)
grps
```

```{r}
table(grps)
```

```{r}
plot(x, col=grps)
```

Hierarchical Clustering is distinct in that the dendrogram (tree figure) can reveal the potential grouping in your data (unlike K-means). 

## Principal Component Analysis (PCA)

PCA is a common and highly useful dimensionality reduction technique used in many fields - particullary bioinformatics.

Here we will analyze some data from the UK on food consumption. 

### Data import

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)

head (x)
```

```{r}
rownames(x) <- x[ ,1]
x <- x[ ,-1]
head(x)
```

```{r}
x <- read.csv(url, row.names = 1)
head(x)
```

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

One conventional plot that can be useful is called a "paris" plot.

```{r}
pairs(x, col=rainbow(10), pch=16)
```

### PCA to the rescue
### Interpreting PCA Results

The main function in base R for PCA is called `prcopm()`

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

The `prcomp()` function returns a list object of our results with five attributes/components.

```{r}
attributes(pca)
```

The two main "results" in here are `pca$x` and `pca$rotation`. The first of these (`pca$x`) contains the scores of the data on the new PC axis - we use these to make our "PCA plot". 

```{r}
pca$x
```

```{r}
library(ggplot2)
library(ggrepel)

#Make a plot of pca$x with PC1 vs PC2
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point() +
  geom_text_repel()
```
We map the relationship between variables in terms of the major principal components.  

The second major result is contained the `pca$rotation` object or component. Let's plot this to see what PCA is picking up...

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```
Fresh potatoes and soft drink have large positive loadings push N. Ireland toward the right side of PCA plot. This indicates that N. Ireland consumes more fresh potatoes and soft drink compared to other countries. Also, fresh fruit and alcoholic drinks have large negative loadings push Scotland, England, and Walves toward the left side of PCA plot. This indicates that Scotland, England, and Walves consume more fresh potatoes and soft drink compared to N.Ireland. 

